Students' Alcohol Consumption Impacts on Academic Results on the Example of a Secondary Scool Maths Class in Portugal

### 2016-10-15 MSproject_Submission.Rmd
Update and rename knit_MSproject.Rmd to MSproject_Submission.Rmd
Add files via upload
Final Submission document includes final layout style changes and missing references

Content:

@@

**Figure 1. General Information**
```{r b, echo=FALSE}
grid.arrange(pie_gender, pie_address, ncol=2)
```

@@

**Figure 2. Family**
```{r c, echo=FALSE}
grid.arrange(pie_guardian, pie_fjob, pie_mjob, widths = c(1, 1, 2), layout_matrix = rbind(c(1,1,1), c(2,2,3)))
```

@@

**Figure 3. Education**
```{r d, echo=FALSE}
grid.arrange(bar_studytime, bar_school, ncol=2)

@@

@@

**Figure 4. Past Class Failures**
```{r e, echo=FALSE}
pie_failures
```

@@

## 3. Hypothesis Testing

As aforementioned, this paper looks into factors influencing academic performance. Our expectation is to see a high correlation between family background and social activities on students' performances. In particular, we are interested to see how alcohol consumption is impacting students' grades. Our basic assumption is that alcohol consumption has a negative impact on students grade. Thus, the Null Hypothesis (H0) to be checked assumes that alcohol consumption does not have an impact on academic performance. To test this hypothesis, multivariate regression analysis will be applied on the 395 samples of students described above. This will allow us to see the significant factors that contribute to academic performance, and check whether or not alcohol consumption is one of them. In this regression, grade is assigned as the dependent variable, and all other variables are treated as independent variables. If alcohol appeared as one of the significant variables in the regression, the null hypothesis will be rejected, inferring that alcohol **is** one of the variables related to academic performance.

@@

## 4. Correlation Analysis and Regression

Prior to regression, there are several items that need to be inspected :

@@

#### a. Distribution of Response Variable

```{r f, echo=FALSE}
sm[,"Score"] <- (sm[,"G1"] + sm[,"G2"] + sm[,"G3"])/3

@@

summary(sm[,"Score"])
```

As seen from the graph, the distribution is slightly right-skewed, but very close to normal distribution. The deviation from the normal distribution could be due to a small sample size. Thus, applying a multivariate regression method is appropriate to test the hypothesis, and no variable transformation is required.

#### b. Correlation Analysis

After performing correlation of all explanatory variables against grade (response variable), four variables appear to have high a correlation to grade :

@@

**4) Frequency of going out with friends**
```{r i1, echo=FALSE}
cor.test(sm$Score,sm$goout)
```

The correlation between the frequency of going out with friends and grade is -0.154. With 95% confidence interval and a p-value of 0.002073, we can reject the null hypothesis, which states that there is no correlation between these two variables, and conclude the frequency of going out with friends are negatively correlated with grades. Students who go out more often are more likely to have lower grades. In the following section, regression analysis will be performed to identify a fitting model which explains the students' academic performances and influencing factors.

**Correlation between Alcohol Consumption and Students' Performances**
```{r i2, echo=FALSE}
cor.test(sm$Score,sm$Walc)
```

As shown above, alcohol consumption does not have a significant correlation to students' grades. The test result shows a correlation of -0.08. With 95% confidence interval and a p-value of 8%, we can **not** reject the null hypothesis, which states that there is no correlation between these two variables. Therefore, we can conclude alcohol consumption does not have a significant impact on students' grades, which will also be verified in the following multivariate regression analysis.

#### c. Regression

After assessing the correlations between all the variables, multivariate regression will be applied. Firstly, all variables are put in the regression model. Secondly, the variable with the highest p-value is eliminated from the model. This iterative process is repeated until a model with a satisfying R2 value can be derived, which consists of significant explanatory variables.

Initial Regression Model :

@@

**Figure 5. Final Regression Model**

@@

stargazer(reg24, type = "text", digits = 2)
```

@@

**Figure 6. Residual Plot**

```{r a1, echo=TRUE}
# residual plot
fit_res <-  data.frame(sm, fitted.value = fitted(reg24), residual = resid(reg24))

@@

####Grades and Alcohol Consumption

Both the correlation test and regression model show that alcohol consumption does not directly affect students' grades. However, examining the correlation between alcohol consumption and the frequency of going out as well as students' study time, which have shown to have an impact on students' grades, demonstrates a high correlation.

@@

## 5. Conclusion

The study done by Keng and Huffman (2010) outlined that alcohol abuse has a negative impact labor market outcomes. However, this correlation could not be replicated for secondary school students: the result of the analysis show no direct correlation between alcohol consumption and secondary school students' academic performance. Alcohol consumption amongst secondary school students is highly correlated to two factors, which do in fact impact students academic performance, namely frequency of going out with friends as well as study time. Further research needs to be done to identify reasons for discrepancy between both studies, which could be due to a variety of different factors. Firstly, the values given in the data (from 1 - very low to 5- very high) do not indicate whether there is abusive alcohol consumption and leave room for personal interpretation. Furthermore, small sample size and age restriction for drinking alcohol could also impact the result. It could also be researched whether the alcohol's impact on people's concentration and performance takes time to show effects and thus would not necessarily impact students' grades at this stage. In conclusion, a more specific study needs to be designed to investigate the impact of these factors and other aspects to draw further conclusions on alcohol consumption in relation to academic achievements.

@@

@@

## 7. Appendices

### Appendix A
Attribute Information:

**Attributes in the Dataset:** 
1 school - student's school (binary: 'GP' - Gabriel Pereira or 'MS' - Mousinho da Silveira)

@@

**Grades Observed in the Dataset:**
31 G1 - first period grade (numeric: from 0 to 20)

>>>>

**Content**:

@@

```{r b, echo=FALSE}
grid.arrange(pie_gender, pie_address, ncol=2)
```

*Figure 1. General Information*

@@

```{r c, echo=FALSE}
grid.arrange(pie_guardian, pie_fjob, pie_mjob, widths = c(1, 1, 2), layout_matrix = rbind(c(1,1,1), c(2,2,3)))
```

*Figure 2. Family*

@@

```{r d, echo=FALSE}
grid.arrange(bar_studytime, bar_school, ncol=2)

@@

*Figure 3. Education*

@@

```{r e, echo=FALSE}
pie_failures
```

*Figure 4. Past Class Failures*

@@

## 3. Hypothesis Testing
As aforementioned, this paper looks into factors influencing academic performance. Our expectation is to see a high correlation between family background and social activities on students' performances. In particular, we are interested to see how alcohol consumption is impacting students' grades. Our basic assumption is that alcohol consumption has a negative impact on students grade. Thus, the Null Hypothesis (H0) to be checked assumes that alcohol consumption does not have an impact on academic performance. To test this hypothesis, multivariate regression analysis will be applied on the 395 samples of students described above. This will allow us to see the significant factors that contribute to academic performance, and check whether or not alcohol consumption is one of them. In this regression, grade is assigned as the dependent variable, and all other variables are treated as independent variables. If alcohol appeared as one of the significant variables in the regression, the null hypothesis will be rejected, inferring that alcohol **is** one of the variables related to academic performance.

@@

## 4. Correlation Analysis and Regression
Prior to regression, there are several items that need to be inspected :

@@

#### a. Distribution of Response Variable
```{r f, echo=FALSE}
sm[,"Score"] <- (sm[,"G1"] + sm[,"G2"] + sm[,"G3"])/3

@@

summary(sm[,"Score"])
```
As seen from the graph, the distribution is slightly right-skewed, but very close to normal distribution. The deviation from the normal distribution could be due to a small sample size. Thus, applying a multivariate regression method is appropriate to test the hypothesis, and no variable transformation is required.

#### b. Correlation Analysis
After performing correlation of all explanatory variables against grade (response variable), four variables appear to have high a correlation to grade :

@@

**4) Frequency of going out with friends**
```{r i1, echo=FALSE}
cor.test(sm$Score,sm$goout)
```
The correlation between the frequency of going out with friends and grade is -0.154. With 95% confidence interval and a p-value of 0.002073, we can reject the null hypothesis, which states that there is no correlation between these two variables, and conclude the frequency of going out with friends are negatively correlated with grades. Students who go out more often are more likely to have lower grades. In the following section, regression analysis will be performed to identify a fitting model which explains the students' academic performances and influencing factors.

**Correlation between Alcohol Consumption and Students' Performances**
```{r i2, echo=FALSE}
cor.test(sm$Score,sm$Walc)
```
As shown above, alcohol consumption does not have a significant correlation to students' grades. The test result shows a correlation of -0.08. With 95% confidence interval and a p-value of 8%, we can **not** reject the null hypothesis, which states that there is no correlation between these two variables. Therefore, we can conclude alcohol consumption does not have a significant impact on students' grades, which will also be verified in the following multivariate regression analysis.

#### c. Regression
After assessing the correlations between all the variables, multivariate regression will be applied. Firstly, all variables are put in the regression model. Secondly, the variable with the highest p-value is eliminated from the model. This iterative process is repeated until a model with a satisfying R2 value can be derived, which consists of significant explanatory variables.

**Initial Regression Model:**

@@

**Final Regression Model**

@@

stargazer(reg24, type = "text", digits = 2)
```
*Figure 5. Final Regression Model*

@@

```{r a1, echo=FALSE}
# residual plot
fit_res <-  data.frame(sm, fitted.value = fitted(reg24), residual = resid(reg24))

@@

*Figure 6. Residual Plot*

####Grades and Alcohol Consumption
Both the correlation test and regression model show that alcohol consumption does not directly affect students' grades. However, examining the correlation between alcohol consumption and the frequency of going out as well as students' study time, which have shown to have an impact on students' grades, demonstrates a high correlation.

@@

## 5. Conclusion
The study done by Keng and Huffman (2010) outlined that alcohol abuse has a negative impact labor market outcomes. However, this correlation could not be replicated for secondary school students: the result of the analysis show no direct correlation between alcohol consumption and secondary school students' academic performance. Alcohol consumption amongst secondary school students is highly correlated to two factors, which do in fact impact students academic performance, namely frequency of going out with friends as well as study time. Further research needs to be done to identify reasons for discrepancy between both studies, which could be due to a variety of different factors. Firstly, the values given in the data (from 1 - very low to 5- very high) do not indicate whether there is abusive alcohol consumption and leave room for personal interpretation. Furthermore, small sample size and age restriction for drinking alcohol could also impact the result. It could also be researched whether the alcohol's impact on people's concentration and performance takes time to show effects and thus would not necessarily impact students' grades at this stage. In conclusion, a more specific study needs to be designed to investigate the impact of these factors and other aspects to draw further conclusions on alcohol consumption in relation to academic achievements.

@@

Keng, S.H. and Huffman, W.E., 2010. Binge drinking and labor market success: a longitudinal study on young people. Journal of Population Economics, 23(1), pp.303-322.

OECD (2013). Education at a Glance 2013. Available at https://www.oecd.org/edu/Portugal_EAG2013%20Country%20Note.pdf. [Accessed 11.10.2016]

Statista (2016). Countries with the highest per capita consumption of alcohol in 2015 (in liters of pure alcohol per year). Available at https://www.statista.com/statistics/280139/countries-with-highest-per-capita-consumption-of-alcohol/ [Accessed 11.10.2016]

@@

## 7. Appendix

**Attributes in the Dataset:**

1 school - student's school (binary: 'GP' - Gabriel Pereira or 'MS' - Mousinho da Silveira)

@@

**Grades Observed in the Dataset:**

31 G1 - first period grade (numeric: from 0 to 20)
